---
title: "Hello my name is <yournamehere>"
author: "Patrick Sadil"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<!-- edit name1 and name2 in the YAML above -->

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
# knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse) # load tidyverse package
```



```{r}
nodes <- DiagrammeR::create_node_df(
  n = 3,
  label = c("zero","low","high"))

DiagrammeR::create_graph(nodes_df = nodes) %>%
  DiagrammeR::render_graph(layout = "tree")
```


---

# Keep your goals in mind

- The best backup and version control systems are the ones that you use continually.

-- 

 - If you're satisfied with Box, Google Drive, USB drives, etc, then keep at it.
 
-- 

- I think that `git` encourages good practices and so using it has the potential to make you a better researcher, teacher, writer.

--

- My goals are to 

-- 

 1) Show you how `git` (with GitHub) can replace your current method
 
 -- 
 
 2) Overview a few ways in which it'll make you better


--- 

# `git` does the version control

- Version Control

-- 

 - Controls the versions
 
--

- Version

-- 

 - For a software engineer, this is 1.0.23
 
-- 

 - For a researcher, a version often means a particular output
 
  - Each experimental pilot
  - The code that made a graph six weeks ago
  - All of your analyses prior to a poster vs. prior to a publication
  
-- 

- As things change, you want to know

 - the exact state of a project at a certain time (reproducible!)
 - *why* it was in that state

???

There are dozens of comics on version control.

---

# Initialize a repository with `git init`

```{bash}

```



---

# Resources I consult regularly

- [Git Pro Book](https://git-scm.com/book/en/v2)
- Repositories of people I trust (e.g., [tidyverse packages](https://github.com/tidyverse))
